---
title: "V2 TabularData Univariate Class Exercises"
author: "Vitalii Zakhozhyi"
output: pdf_document
---

**Calling data and preparing it for exercises**

First, we need to call data and prepare it for the excercises (cheking the structure, types of veriable, etc)

```{r getData, eval=FALSE}
link='https://github.com/EvansDataScience/VisualAnalytics_2_tabularData/raw/master/data/eduwa.rda'
load(file=url(link))
```

```{r dataSTR, eval=FALSE}
str(eduwa,width = 100,strict.width='cut')
```

Then, we build the table of frequences, which we will be working with further

```{r freqTab, eval=FALSE}
absoluteT=table(eduwa$LocaleType,exclude = 'nothing')
names(absoluteT)[5]='Unknown'
ToPlot=prop.table(absoluteT)*100
ToPlot
```

**Exercise 1: Use ggplot to show a bar plot ordered by share size.**

Excercise 1 solution:

```{r barplot_gg_ex, eval=FALSE}
library(ggplot2)
library(scales)

tableFreq=as.data.frame(ToPlot) 
names(tableFreq)=c("locale","pct")
tableFreq_Ordered=tableFreq[order(tableFreq$pct),] #ascending ordering in ggplot

barplot = ggplot(tableFreq_Ordered, aes(x = locale, y = pct)) + #ggplot, data, aesthetics (mapping x, y)
            geom_bar(fill ="gray", stat = 'identity') + #geometry barplot
            labs(title='Are all locations getting a fair share of public schools in WA?', #label title
                 x ='Type of location', #label x axis
                 y = 'Share of public schools', #label y axis
                 caption = 'Source: US Department of Education') + #label caption
            geom_hline(yintercept = 25, # creating reference line as horisontal line with the next attributes
                       linetype="dashed", #line type
                       size=1.5,  #line width
                       alpha=0.5) + #transparency
            coord_flip() +  #flipping the plot
            scale_y_continuous(breaks=c(0,10, 25,40,50), #playing with scales and labels of y axis
                               limits = c(0, 50),
                               labels=scales::unit_format(suffix = '%')) +
            scale_x_discrete(limits=tableFreq_Ordered$locale) + #changing the indexes and the limits/categories 
            theme(panel.background = element_rect(fill = "white", colour = "grey50"), #changing a theme
                  plot.caption = element_text(hjust = 1), #location of caption source
                  plot.title = element_text(hjust = 0.5)) + #location of title
            geom_text(aes(y = pct , #addind a text with indication of share to each bar
                          label = paste0(round(pct,2), '%')), #paste0 means that we combine 2 elements but without space between them
                          hjust= -0.1, #location of the text (<0 - outside the bar, >0 - inside the bar); vjust - for vertical, hjust - for horisontal bars 
                          size = 3)
barplot
```



**Exercise 2: Complete adding the elements missing in the lolliplot (with a reference line).**

Excercise 2 solution:

```{r lolli_gg_ex, eval=FALSE}
tableFreq$gap=tableFreq$pct-25 # creating new variable "gap" to show gaps
tableFreq$PositiveGap=ifelse(tableFreq$gap>0,'Above 25% baseline', 'Below 25% baseline') # creating variable to show gaps with color

lolliplot = ggplot(tableFreq, aes(x = locale,
                                  y = gap,
                                  color = PositiveGap,
                                  label = round(gap,2))) + #ggplot, data, aesthetics (mapping x, y), color = using gap, label  - adding labels with gap value
            geom_segment(aes(y = 0, #creating vertical lines/bars; aesthetics (assigning y start of line)
                             x = locale, #assigning x start of line
                             yend = gap, #assigning y end of line
                             xend = locale), color = "grey50") + #assigning x end of line; changing color
            geom_point() + #adding points to the plot - creating lolliplot
            labs(title='Are all locations getting a fair share of public schools in WA?', #label title
                 x ='Type of location', #label x axis
                 y = 'Difference from 25% baseline', #label y axis
                 caption = 'Source: US Department of Education') + #label caption
            geom_text(nudge_x=0.3, show.legend = FALSE) +  # annotating and moving the text on the horizontal
            scale_x_discrete(limits=tableFreq[order(-tableFreq$gap),]$locale) + #ordering and rescaling
            theme(plot.caption = element_text(hjust = 1), #changing a theme; location of caption source
                  plot.title = element_text(hjust = 0.5), #location of title
                  legend.title = element_blank()) #delete legend title 
lolliplot
```




**Exercise 3: Make sure our box plot follows the same design approach and include all the elements as in the bar plot for nominal data.**

Excercise 3 solution:

```{r boxplot_gg_ex, eval=FALSE}
boxplot = ggplot(eduwa, aes(y=as.numeric(High.Grade))) + #data, aesthetics (mapping only y for this type of plot), High.Grade is ordinal variable, but boxplot expects from us a numerical var, so we transform High.Grade into numerical by using = as.numeric()
            geom_boxplot() + #boxplot geometry
            coord_flip() + #flipping the plot
            scale_y_continuous(labels=levels(eduwa$High.Grade),breaks=1:15) + #assigning new labels and changing scaling
            labs(title= 'Concentration of schools in WA by the highest grade they offer', #label title
                 y = 'Grade level', #label y axis
                 caption = 'Source: US Department of Education') + #label caption
            theme(panel.background = element_rect(colour = "grey50"), #change of background color
                  plot.caption = element_text(hjust = 1), #caption location
                  plot.title = element_text(hjust = 0.5), #title location
                  axis.text.y = element_blank(), #remove value lables
                  axis.ticks.y = element_blank()) #remove ticks
boxplot
```




**Exercise 4: Make a histogram for the variable FREE LUNCH, and make sure it has all the right elements, and get rid of unnecessary elements.**

Excercise 4 solution:

```{r freelunch, eval=FALSE}
summary(eduwa$Free.Lunch)
eduwa_freelunch=eduwa[complete.cases(eduwa$Free.Lunch),]
```


```{r hist_freelunch_gg_ex, eval=FALSE}
hist_freelunch = ggplot(eduwa_freelunch,aes(x = Free.Lunch)) + 
                    geom_histogram(color="black", fill="lightblue", binwidth = 150, boundary=0) + #changing color, filling, and width
                    stat_bin(binwidth = 150, aes(label=..count..), ##putting numbers of values in each bin above bins using count
                             geom = "text", size = 3, boundary = 0, vjust=-0.5) + 
                    scale_x_continuous(breaks = seq(0, 1800, by = 150)) + #x axis breaks
                    labs(title= 'Distribution of schools by the number of free lunches offered there', #label title
                         y = 'Number of Schools', #label y axis
                         x = 'Number of Students with a Free Lunch', #label x axis
                         caption = 'Source: US Department of Education') + #label caption
                    theme(plot.caption = element_text(hjust = 1), #caption location
                          plot.title = element_text(hjust = 0.5), #title location
                          axis.title.y = element_text(size = 10), #size of the axis name text 
                          axis.title.x = element_text(size = 10)) #size of the axis name text 
hist_freelunch
```




**Exercise 5: Do some research and make a histogram and a density plot using ggplot for the variable we just used above.**

Excercise 5 solution:

```{r STRmean, eval=FALSE}
eduwa_STR=eduwa[complete.cases(eduwa$Student.Teacher.Ratio),]
eduwa_STR_mini = select(eduwa_STR, "County", "Student.Teacher.Ratio")
STRmean = tapply(eduwa_STR_mini$Student.Teacher.Ratio, 
                 eduwa_STR_mini$County,
                 mean)
STRdf = as.data.frame(STRmean)
summary(STRdf)
```

```{r hist_STRmean_gg_ex, eval=FALSE}
hist_STRmean = ggplot(STRdf,aes(x = STRmean)) + 
                    geom_histogram(color="black", fill="lightblue", binwidth = 2, boundary=0) + #changing color, filling, and width
                    scale_y_continuous(breaks = seq(0, 20, by = 2)) + #y axis breaks
                    scale_x_continuous(breaks = seq(10, 24, by = 2)) + #x axis breaks
                    geom_vline(aes(xintercept = mean(STRmean), # creating reference line as vertical line (mean of mean STRs of counties) with the next attributes
                               color = "Mean"), #line color
                               linetype="dashed", #line type
                               size=1.3,  #line width
                               alpha=0.8) + #transparency
                    labs(title= "Distribution of Student-Teacher Ratio Counties' Means", #label title
                         y = 'Frequency', #label y axis
                         x = 'Mean Student/Teacher Ratio', #label x axis
                         caption = 'Source: US Department of Education') + #label caption
                    theme(plot.caption = element_text(hjust = 1), #caption location
                          plot.title = element_text(hjust = 0.5), #title location
                          axis.title.y = element_text(size = 10), #size of the axis name text 
                          axis.title.x = element_text(size = 10), #size of the axis name text 
                          legend.title = element_blank()) #delete name of the legend

hist_STRmean
```

```{r dens_STRmean_gg_ex, eval=FALSE}
dens_STRmean = ggplot(STRdf,aes(x = STRmean)) + #data and aesthetics, mapping x
                  geom_density(color="black") + #changing color
                  scale_x_continuous(breaks = seq(10, 24, by = 1)) + #x axis breaks
                  geom_vline(aes(xintercept = mean(STRmean), # creating reference line as vertical line (mean of mean STRs of counties) with the next attributes
                             color = "Mean"), #line color
                             linetype="dashed", #line type
                             size=1.3,  #line width
                             alpha=0.8, #transparency
                             show.legend = TRUE) + #show legend 
                  geom_vline(aes(xintercept = median(STRmean), # creating reference line as vertical line (median of mean STRs of counties) with the next attributes
                             color = "Median"), #line color
                             linetype="dashed", #line type
                             size=1.3,  #line width
                             alpha=0.8, #transparency
                             show.legend = TRUE) + #show legend                
                  labs(title= 'Density of Student-Teacher Ratio', #label title
                       y = 'Density', #label y axis
                       x = 'Mean Student/Teacher Ratio', #label x axis
                       caption = 'Source: US Department of Education') + #label caption
                  theme(plot.caption = element_text(hjust = 1), #caption location
                        plot.title = element_text(hjust = 0.5), #title location
                        axis.title.y = element_text(size = 10), #size of the axis name text 
                        axis.title.x = element_text(size = 10), #size of the axis name text 
                        legend.title = element_blank()) #delete name of the legend
dens_STRmean
```
